---
title: Fisher's Analysis of Iris Data
thanks: 
  - ref: T1 
    text: Based on the article "The Use of Multiple Measurements in Taxonomic Problems" by R. A. Fisher (1936)
runtitle: Iris Data 
author:
  - name: Qian 
    surname: Zhao
    email: qzhao1@stanford.edu
    label: e1
    addressLabel: A


affiliation:
  - label: A
    name: Department of Biomedical Data Science, Stanford University
    authorsLabels: e1
abstract: |
  I use an analysis of the Iris data set to illustrate how to use the "rticles" package to create a reproducible manuscript. 
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - reproducible manuscript
  - iris data

predefined-theoremstyle: true # use in section Environments for Axiom, Theorem, etc
bibliography: ims.bib
biblio-style: imsart-nameyear # alternative: imsart-number
output:
  rticles::ims_article:
    journal: aoas # aap, aoas, aop, aos, sts. See documentation
    toc: false # Please use for articles with 50 pages and more

header-includes: 
 - \usepackage{caption} 
 - \usepackage{subcaption}
---



```{r load_data, echo = F}
data(iris) 
```

# The iris data
The Iris data, collected by Dr. E. Anderson, contains measurements of the flowers of fifty plants each of the two species \textit{Iris setosa} and \textit{I.versicolor}. Figure \ref{fig:iris} shows pictures of the two species. The data includes four measurements: sepal length, sepal width, petal length, and petal width. A few rows of the data are shown in Table 

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=4cm]{setosa}
         \caption{\textit{Iris setosa}}
     \end{subfigure}
     \begin{subfigure}[b]{0.4\textwidth}
         \centering
         \includegraphics[width=4cm]{versicolor.jpeg}
         \caption{\textit{I.versicolor}}
     \end{subfigure}
        \caption{Two iris species}
        \label{fig:iris}
\end{figure}

```{r iris_data, echo = F}
tab <- kableExtra::kbl(iris[1:5, ], caption = "First few rows in the iris data", row.names = F,  vline = "", linesep = "")
kableExtra::kable_styling(tab, latex_options = "hold_position")
```

# Fisher linear discriminant analysis

In a 1936 article, @fisher1936 considered the question: what linear function of the four measurements
\begin{equation}
X = \lambda_1 x_1 + \lambda_2 x_2 + \lambda_3 x_3 + \lambda_4 x_4
\end{equation}
maximizes the \textit{ratio} of the difference between the means to the standard deviation within species? 

The observed means and their differences are shown in Table \ref{tab:iris_mean}. We can also compute the sum of squares and products of deviation from specific means of each species (Table \ref{tab:compute_var}). 


```{r compute_mean, echo = F}
iris_means <- dplyr::tibble(
  Variable = c("Sepal length", "Sepal width", 
               "Petal length", "Petal Width"),
  Versicolor = colMeans(subset(iris, Species == "versicolor")[,-5]),
  Setosa = colMeans(subset(iris, Species == "setosa")[,-5]),
  Difference = Versicolor - Setosa
)
```


```{r iris_mean, echo = F}
tab2 <-  kableExtra::kbl(iris_means, caption = "Observed means for two species and their difference (cm.)", row.names = F, format = "latex", vline = "", linesep = "") 
kableExtra::kable_styling(tab2, latex_options = "hold_position")
```

```{r compute_var, echo = F}
iris_centered <- rbind( scale(subset(iris, Species == "setosa")[,-5], scale = F ),
       scale(subset(iris, Species == "versicolor")[,-5], scale = F )) 

cov <- t(iris_centered) %*% iris_centered
row.names(cov) <- c("Sepal length", "Sepal width",  "Petal length", "Petal Width")
colnames(cov)<- c("Sepal length", "Sepal width",  "Petal length", "Petal Width")

tab3 <- kableExtra::kbl(cov, 
             caption = "Sums of squares and products of four measurements, within species (cm.2)", 
             row.names = T, format = "latex",  vline = "", linesep = "")
kableExtra::kable_styling(tab3, latex_options = "hold_position")
```

The linear combination that maximizes $D^2/S$, where
\begin{equation}
D = \lambda_1 d_1 +\lambda_2 d_2+\lambda_3 d_3+\lambda_4 d_4,
\end{equation}
where $d_i$ are the differences between the species means and 
\begin{equation}
S = \sum_{p=1}^4\sum_{q=1}^4 \lambda_p\lambda_qS_{pq},
\end{equation}
is the solution to a set of linear equations 
\begin{equation}
\begin{cases}
S_{11}\lambda_1 + S_{12}\lambda_2 + S_{13}\lambda_3 + S_{14}\lambda_4 = d_1, \\
S_{21}\lambda_1 + S_{22}\lambda_2 + S_{23}\lambda_3 + S_{24}\lambda_4 = d_2, \\ 
S_{31}\lambda_1 + S_{32}\lambda_2 + S_{33}\lambda_3 + S_{34}\lambda_4 = d_3, \\ 
S_{41}\lambda_1 + S_{42}\lambda_2 + S_{43}\lambda_3 + S_{44}\lambda_4 = d_4. \\ 
\end{cases}
\end{equation}

```{r, echo = F}
lambda = solve(cov) %*% iris_means$Difference
```

For the iris data, the solution is `r paste0("(",paste(round(lambda,2), collapse=","),")")`.


